Canonical rings of $\bq$-divisors on Minimal Rational Surfaces

### 2015-08-20 surfaces-write-up/surface-write-up.tex
answered peters pre-arxiv todos

\begin{lem}
\label{lem:composite-map}
Let $X$ be an arbitrary scheme and let $D = \sum_{i=0}^{n}\alpha_i D_i\in \bq \otimes_\bz \di X$ where $D_i = V(f_i)$. Suppose we have a
surjection $\phi\colon \bk[w_1,\ldots, w_r] \rightarrow R_D$ given by $w_i
\mapsto p_i(f_0, \ldots, f_n),$ where $p_i$ is a monomial in $f_0,\ldots, f_n$.
Let $\bida_0, \ldots, \bida_n, \bidb_0, \ldots, \bidb_n \in \bz_{\geq 0}.$
Then, define
\begin{align*}
	\Sigma = \langle u^d z_0^{c_0} \cdots z_n^{c_n} : c_i \geq -d \alpha_i, \sum_{i=0}^{n} \bida_i c_i = \sum_{i=0}^{n} \bidb_i c_i = 0 \rangle. 
\end{align*}
In this case, we can factor $\phi$ as a composition of $\chi$ and $\psi$ defined by
\[
\begin{tikzcd}
\bk[w_1,\ldots, w_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
w_i \arrow[mapsto]{r} & u^{d_i}z_0^{c_{i0}} \cdots z_n^{c_{in}} \arrow[mapsto]{r} & u^{d_i}f_0^{c_{i0}} \cdots f_n^{c_{in}}.
\end{tikzcd}
\]
Assuming $\chi$ is surjective, the minimal degree of generation of $\ker \phi$
is at most the maximum of the minimal degree of generation of $\ker \chi$ and
the minimal degree of generation of $\ker \psi$.
\end{lem}
\begin{proof}
First, we show there is an exact sequence
\[
\begin{tikzcd}
0 \ar {r} & \ker \chi \ar{r}{\rho} & \ker \phi \ar {r}{\tau} & \ker \psi \ar r & 0.
\end{tikzcd}
\]
To see this, note that $\rho$ is injective and $\tau$ is surjective\ because
$\phi = \chi \circ \psi$. Additionally, $\tau \circ \rho = 0$ by definition of
the kernel of $\chi$. Finally, if $a \in\ker \tau$, then by definition $\chi(a)
= 0,$ so $a \in \im \rho$.

This shows that lifts of generators of $\ker \psi$ together with images of
generators of $\ker \chi$ generate all of $\ker \phi,$ as desired. 
\end{proof}\todo{Peter: Maybe we can get rid of most of this proof, and replace it with: ``The proof follows by examining the exact sequence $0 \to \ker\chi \to \ker \phi \to \ker \psi \to 0$.''  I don't think it is necessary to prove why this is an exact sequence, as it is taught in a first commutative algebra course.}

\begin{lem}
\label{lem:bound-ker-chi}
Retaining the notation of Lemma ~\ref{lem:composite-map}, if $\Sigma$ has
extremal rays $e_0,\ldots, e_t$ in degrees $d_0, \ldots, d_t$ then $\ker \chi$
is generated in degrees at most $2(\sum_{i=0}^{t}d_i-1)$.
\end{lem}
\begin{proof}
Since $e_0, \ldots, e_t$ are extremal rays, Proposition
~\ref{prop:cone-generation} tells us that every element $\sigma \in \Sigma$ can be written
in a canonical form $\lambda + \sum_{i=0}^{t} \zeta_i e_i$ where all $\zeta_i \in \mathbb{Z}_{\ge 0}$.  Let
$\lambda_0 :=
0,\lambda_1, \ldots, \lambda_r$ be all elements of $\Sigma$ which
can be
written in the form $\lambda_j = \sum_{i=0}^{t}s_i e_i$ with $0
\leq s_i < 1.$ Then, for any $1 \leq j \leq k \leq r,$ we can write $\lambda_j + \lambda_k$ in
the above canonical form, yielding a (possibly trivial) relation in degree at most $\deg \lambda_j
+ \deg \lambda_k\leq 2 \cdot \left( \sum_{i=0}^{t}d_i -1 \right).$
Furthermore, these relations generate all relations, as one can apply a
sequence of these relations to put any $\sigma \in \Sigma$ into canonical form
$\sigma = \lambda + \sum_{i=0}^{t}\zeta_i e_i$ from proposition Proposition
~\ref{prop:cone-generation}.
%as in Equation
%~\eqref{eqn:sigma-canonical-form}.
\end{proof}
\begin{lem}
\label{lem:proj-relations-psi}
Let $D = \sum_{i=0}^{n} \alpha_i D_i \in \bq \otimes_\bz \di \bp^m,$ where
$\deg D_i = \bida_i$, $\alpha_i = \frac{c_i}{k_i}\in \bq$, and
$\ell_i = \lcm_{j \neq i} (k_j)$. Define $\Sigma$ as in Equation
~\eqref{eqn:Sigma-defn} and $\psi$ as ~\eqref{eqn:factor-through-semigroup-ring}. Then, $\ker \psi$ is generated in degrees at most
\begin{align}
\label{eqn:proj-relation-degree}
	\frac{\max_{0\le i \le n}(\bida_i)}{\deg(D)} +  \sum_{i=0}^n \ell_i a_i.
\end{align}

@@

Furthermore, similarly to the case of projective space, 
we will restrict to the case that $D$ is a divisor
for which both of its bi-degrees are positive \todo{Do we need to specify what we mean by the bi-degrees of a divisor, or will that be apparent to a reader?}. We now justify this
restriction. If either of the bi-degrees of $D$ are negative,
then the canonical ring is concentrated in degree 0. If one of the
bi-degrees is 0, say the first one is 0, then $R_D$ is isomorphic
to $R_{D'}$, where $D' \in \bq \otimes_\bz \di \bp^1$, 
where $D'$ can be written as a sum of divisors whose degrees
are multiples of the second bi-degree of $D$. Since the case of
$\bp^1$ has already been analyzed in \cite{dorney:canonical}, we
are justified in assuming that both bi-degrees of $D$ are positive.

@@

\section{Acknowledgments}
\label{sec:ack}
We are grateful to David Zureick-Brown for suggesting this problem 
and for providing invaluable guidance. We also thank Ken Ono and the
Emory University Number Theory REU for arranging our project and
providing a great environment for mathematical learning and
collaboration.
Finally, we gratefully acknowledge the financial support given by
NSF Grant Award Number 1250467 via the Emory University Number
Theory REU. We deeply appreciate all of the support that has made
our work possible.\todo{Peter: Add acknowledgements to a few more peopled - Robin's dad; was there anyone else?}

>>>>

\begin{lem}
\label{lem:composite-map}
Let $X$ be an arbitrary scheme and let $D = \sum_{i=0}^{n}\alpha_i D_i\in \bq \otimes_\bz \di X$ where $D_i = V(f_i)$. Suppose we have a
surjection $\phi\colon \bk[w_1,\ldots, w_r] \rightarrow R_D$ given by $w_i
\mapsto p_i(f_0, \ldots, f_n),$ where $p_i$ is a monomial in $f_0,\ldots, f_n$.
Let $\bida_0, \ldots, \bida_n, \bidb_0, \ldots, \bidb_n \in \bz_{\geq 0}.$
Then, define
\begin{align*}
	\Sigma = \langle u^d z_0^{c_0} \cdots z_n^{c_n} : c_i \geq -d \alpha_i, \sum_{i=0}^{n} \bida_i c_i = \sum_{i=0}^{n} \bidb_i c_i = 0 \rangle. 
\end{align*}
In this case, we can factor $\phi$ as a composition of $\chi$ and $\psi$ defined by
\[
\begin{tikzcd}
\bk[w_1,\ldots, w_r] \ar {r}{\chi} & \bk[\Sigma] \ar {r}{\psi} & R_D \\
w_i \arrow[mapsto]{r} & u^{d_i}z_0^{c_{i0}} \cdots z_n^{c_{in}} \arrow[mapsto]{r} & u^{d_i}f_0^{c_{i0}} \cdots f_n^{c_{in}}.
\end{tikzcd}
\]
Assuming $\chi$ is surjective, the minimal degree of generation of $\ker \phi$
is at most the maximum of the minimal degree of generation of $\ker \chi$ and
the minimal degree of generation of $\ker \psi$.
\end{lem}
\begin{proof}
First, note that surjectivity of $\chi$ implies we have an exact sequence
\[
\begin{tikzcd}
0 \ar {r} & \ker \chi \ar{r}{\rho} & \ker \phi \ar {r}{\tau} & \ker \psi \ar r & 0.
\end{tikzcd}
\]
This shows that lifts of generators of $\ker \psi$ together with images of
generators of $\ker \chi$ generate all of $\ker \phi,$ as desired. 
\end{proof}
\begin{lem}
\label{lem:bound-ker-chi}
Retaining the notation of Lemma ~\ref{lem:composite-map}, if $\Sigma$ has
extremal rays $e_0,\ldots, e_t$ in degrees $d_0, \ldots, d_t$ then $\ker \chi$
is generated in degrees at most $2(\sum_{i=0}^{t}d_i-1)$.
\end{lem}
\begin{proof}
Since $e_0, \ldots, e_t$ are extremal rays, Proposition
~\ref{prop:cone-generation} implies every element $\sigma \in \Sigma$ can be written
in a canonical form $\lambda + \sum_{i=0}^{t} \zeta_i e_i$ where all $\zeta_i \in \mathbb{Z}_{\ge 0}$.  Let
$\lambda_0 :=
0,\lambda_1, \ldots, \lambda_r$ be all elements of $\Sigma$ which
can be
written in the form $\lambda_j = \sum_{i=0}^{t}s_i e_i$ with $0
\leq s_i < 1.$ Then, for any $1 \leq j \leq k \leq r,$ we can write $\lambda_j + \lambda_k$ in
the above canonical form, yielding a (possibly trivial) relation in degree at most $\deg \lambda_j
+ \deg \lambda_k\leq 2 \cdot \left( \sum_{i=0}^{t}d_i -1 \right).$
Furthermore, these relations generate all relations, as one can apply a
sequence of these relations to put any $\sigma \in \Sigma$ into canonical form
$\sigma = \lambda + \sum_{i=0}^{t}\zeta_i e_i$ from proposition Proposition
~\ref{prop:cone-generation}.
%as in Equation
%~\eqref{eqn:sigma-canonical-form}.
\end{proof}
\begin{lem}
\label{lem:proj-relations-psi}
Let $D = \sum_{i=0}^{n} \alpha_i D_i \in \bq \otimes_\bz \di \bp^m,$ where
$\deg D_i = \bida_i$, $\alpha_i = \frac{c_i}{k_i}\in \bq$, and
$\ell_i = \lcm_{j \neq i} (k_j)$. Define $\Sigma$ as in Equation
~\eqref{eqn:Sigma-defn} and $\psi$ as ~\eqref{eqn:factor-through-semigroup-ring}. Then, $\ker \psi$ is generated in degrees at most
\begin{align}
\label{eqn:proj-relation-degree}
	\frac{\max_{0\le i \le n}(\bida_i)}{\deg(D)} +  \sum_{i=0}^n \ell_i a_i.
\end{align}

@@

Furthermore, similarly to the case of projective space, 
we will restrict to the case that $D$ is a divisor
for which both of its bi-degrees are positive. We now justify this
restriction. If either of the bi-degrees of $D$ are negative,
then the canonical ring is concentrated in degree 0. If one of the
bi-degrees is 0, say the first one is 0, then $R_D$ is isomorphic
to $R_{D'}$, where $D' \in \bq \otimes_\bz \di \bp^1$, 
where $D'$ can be written as a sum of divisors whose degrees
are multiples of the second bi-degree of $D$. Since the case of
$\bp^1$ has already been analyzed in \cite{dorney:canonical}, we
are justified in assuming that both bi-degrees of $D$ are positive.

@@

\section{Acknowledgments}
\label{sec:ack}
We are grateful to David Zureick-Brown for suggesting this problem 
and for providing invaluable guidance. We thank Shou-Wu Zhang and Peter
Landesman for reading this paper and proving helpful suggestions.
We also thank Ken Ono and the
Emory University Number Theory REU for arranging our project and
providing a great environment for mathematical learning and
collaboration.
Finally, we gratefully acknowledge the financial support given by
NSF Grant Award Number 1250467 via the Emory University Number
Theory REU. We deeply appreciate all of the support that has made
our work possible.
